|

### 2020-11-19 paper/v1/broadband.speed.covid.Rmd

During the pandemic, working from home using Information and Communication
Technologies (ICT), whether partially or exclusively, was transformed from a
niche means of accessing work, albeit one that had been on a slow, upward trend,
to a widespread way of life in many countries [@felstead2020homeworking].
**HANNAH, DO WE REALLY NEED A REFERENCE IN THE PREVIOUS SENTENCE?**
The ability to work from home or telecommute meant millions retained their jobs
and, to a varying extent, maintained productivity during periods of strict
lockdown around the world.
<!--OLD TEXT: However, this ability was not evenly distributed socially or spatially, creating
a new type of digital divide -- one of economic resilience and personal safety on
one side and unemployment or essential front-line work on the other.
The ability to telecommute is partly a product of whether the majority of work
tasks in an occupation can be performed independently of being in a particular
location or co-locating with colleagues.-->
However, this ability has not been evenly distributed socially or spatially,
creating a new type of digital divide.
On one side are those who can work from home and have been able to enjoy both
economic resilience and greater personal safety.
On the other side, previously employed individuals have been forced to accept
furlough or redundancy packages unless they are part of the cadre of essential
workers, who are potentially at high risk of infection.
Using the UK as a case study, this paper aims to improve our understanding of
the spatial and social dimensions of this new digital divide by assessing the
resilience of broadband speeds in terms of both quality and reliability of
service, and whether this reinforces or redresses prior digital divisions.
<!--and upload speeds in the UK in order to gain insights into both the quality and
reliability of broadband services across the country and the distribution of
telecommuting at the time when the population were told to work from home if
possible.-->
To do so, we employ unique volunteered geographic data on individual broadband
speed tests and state-of-the-art time-series clustering methods, which enable us
to create clusters of UK local authorities with similar temporal signatures of
experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore whether they are linked with existing
divides and the economic geography of the UK.
**HANNAH, HOW DOES THIS SOUND?**
Our analysis enables us to better understand how telecommuting and
technology intersect at a time of extreme demand, and what lessons this time has
for a future where telecommuting is likely to remain a common means of accessing
work and broadband services, as well as infrastructure, must be fit for purpose.
**LET'S LEAVE IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ**

<!-- Our framework: capability to WFH; HB old PARA2 -->
The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and are
allowed to be performed independently of location or co-location with
colleagues, including supervisors (**insert refs**).
However, successful telecommuting also requires that the quality and
reliability of ICT services, particularly home internet connection speeds,
enable the completion of work tasks with a minimum of delay or interruption.
In reality, the performance of broadband speeds with respect to telecommuters
has never been tested before at scale, as working from home and connecting to
colleagues and workplace resources via broadband has previously been the
purview of a small minority of workers.
This papers understands telecommuting as a function of the quality of ICT
services, particularly home internet connections, and whether work
tasks in an occupation can be performed independently of being in a particular
location or co-locating with colleagues

<!-- How much do we work form home-->
The shift towards telecommuting during various stages of lockdown around the world
has been drastic and, importantly, speculations indicate that the post-Covid
tendency to work from home will be much higher than the pre-Covid one.
A back of the envelope calculation suggests that up to 40% of the working force
could work from home in the UK [@batty2020editorial].
Observational data pointed an even higher share of people in employment in the
UK who worked from home in April $2020$ ($47$%), while the same figure only reached
$5$% the year before [@ons2020; @ons2020lm2019].
Similar figures have been reported for other countries around the world
[@felstead2020homeworking].
For instance, $37$% of the workforce worked from home in Europe in April $2020$ with
countries like Finland reaching $60$% [@eurofound2020].
In the US almost half of the working population worked from home during the same
period because of the pandemic [@brynjolfsson2020covid].
More broadly, a recent estimate indicated that $37$% of all jobs in the US can be
permanently performed entirely from home [@NBERw26948].

<!-- Who works from home-->
There is a consensus that opportunities for working from home especially during
the current pandemic are not equally spread across the workforce.
@NBERw26948 indicated that in the US managers, educators, as well as those
working in computer-related occupations, finance, and law can easily work from
home.
On the contrary, workers in farming, construction, and manufacturing do not have
such opportunities.
Not surprisingly, occupations with opportunities to telecommute are associated
with higher earnings.
This is not the case for the workforce occupied in less footloose occupations as
they tend to be lower-income, non-white, without a university degree,
live in rental accommodation and lack health insurance [@NBERw27085].
Although these figures refer to the US, similar trends can be observed for other
countries.
For example, $75$% of workers with tertiary education worked from home in Europe
during spring $2020$, while the same share for workers with secondary and
primary education dropped to $34$% and $14$% respectively.
Moreover, employees living in cities, women and younger employees were
have worked from home [@eurofound2020].

<!-- ICT infrastructure-->
None of these changes could have happened in the absence of reliable ICT
infrastructure -- both in terms of software and hardware.
But while software innovations are easily diffused across space and
society^[See for example the huge success of videoconferencing apps
[@marks2020zoom].],
the same does not apply for ICT hardware infrastructure such as internet
broadband connectivity.
The literature exemplifies digital divides in terms of internet access and its
quality.
For instance, @riddlesden2014broadband highlighted the broadband divides in the
UK, while the systematic review from @SALEMINK2017360 reinforced our
understanding for the infrastructure quality differences between urban and rural
areas.

<!-- WFH and digital divides-->
Our framework to understand telecommuting as a function of occupations and
quality of ICT infrastructure is aligned with current debates on digital divides.
While the, so-called, first level digital divides are associated with access
and quality of internet connectivity, the second level ones are linked
to the necessary skills to effectively utilise ICT and the internet
[@blank2014dimensions; @van2011internet].
Importantly, the capacity to telecommute, which to a certain
extend is related to the first and second level digital divides, leads to
differentiated outcomes regarding the economic resilience of people and
places against the current pandemic.
And to the extend that the quality of the internet infrastructure and occupation
variation are spatially dependent and clustered in space, the spatial footprint
of telecommuting is of great interest. In a way similar to the third level
digital divide, which
focuses on the differentiated returns of internet use [@stern2009levels;
@van2014digital; @van2015third] places with high rates of telecommuting during
the Covid pandemic illustrate higher economic resilience against the current pandemic.
<!-- Econ resilience-->
This is aligned with the regional economic resilience literature, which
underlines the differentiated capacity of cities and regions to escape or
recover from economic crises [@martin2012regional; @kitsos2018economic].

<!-- WFH broader consequences-->
The long-term effects of such drastic changes in telecommuting and attitudes
towards working from home are difficult to predict. Nevertheless,
they span through various aspects of economy and society: from changes to
transportation planning due to altered commuting patters, to changes in land
use and urban planning to accommodate people who work from home
[@BUDNITZ2020102713]; and from productivity and innovation changes, to changes
in agglomeration externalities and the attraction of large cities [@econobs]
just to name a few. This paper is positioned to support endeavours in
understanding the effects of increased telecommuting by
exposing the spatial and social dimensions of telecommuting including the
resilience of broadband speeds in terms of both quality and reliability of
service, and whether this reinforces or redresses prior digital divisions.
**HANNAH, HOW DOES THIS SOUND**

The structure of this paper is as follows.
...

@@

## broadband studies, divides, broadband tech stuff

- PARA about Contention??

<!-- WFH and infrastructure, HB old PARA2 -->
The infrastructural demands of working from home on internet services have been
minor compared to the demands of leisure users, such that the broadband
performance offered by different Internet Service Providers (ISPs) has been
benchmarked according to download speeds during the evening 'primetime',
when video streaming services are at their peak, rather than during the working
day (**OfCom, 2017**).
Yet the pandemic has fundamentally changed not only how many people work from
home, but also their technical requirements when doing so.
Whilst it may have been common in the past for telecommuters to complete
solitary work tasks at home, but still visit their workplace for meetings
(**ref?**), during the pandemic the replacement of face-to-face contact with
bandwidth-intensive video conference calling was seen as essential to
suppressing the spread of infection.
This type of broadband use can particularly affect upload
speeds **ref?**.
This paper offers a framework to assess the quality and reliability of upload
speeds by time of day and day of the week across local authorities in the UK
during the time when the population were told to work from home if at all
possible.

@@

@@

In this analysis, the terms ‘teleworking’ and ‘working from home’ are used interchangeably. An appropriate definition of teleworking is ‘the remote
provision of labour that would otherwise be carried out within company premises’ (European Commission, 2020b). In practice, during the COVID crisis,
most such work was carried out in the homes of individual employees rather than any other location.
[@eurofound2020]

@@

<!-- HB old PARA3 -->
The experience of upload speeds is not the same as the maximum speeds offered by an ISP or possible speeds over a particular type and length of connection. Therefore, we use volunteered geographic information in the form of speed-checks run by users to test their experienced broadband speeds, upload and download. Meanwhile, the quality and reliability of upload speeds, like ICT more generally, vary in time and space due to both supply and demand-side influences, and can be measured in a number of ways. These include: a) mean, experienced connection speed, b) standard deviation or the amount of fluctuation from the mean, and c) the variation in speeds at particular times of day when working from home is more likely to take place. We take account of all three measurements in order to describe upload speeds as fully as possible. We start by calculating the mean upload speed for every local government district in the UK for each hour of each weekday, excluding midnight to 6:00, weekends and bank holidays. We then cluster the local authorities by these temporal profiles, allowing us to identify patterns and describe the overall means, standard deviations, and other relevant statistics for each spatial cluster. Thus we aim to answer the first part of our research question: How resilient are broadband speeds as experienced in different parts of the UK during a time of extreme demand?

@@

# Conclusions

>>>>

During the pandemic, working from home using digital technologies, whether partially 
or exclusively, was transformed from a niche means of accessing work, albeit one that 
had been on a slow, upward trend, to a widespread way of life in many countries.
The ability to work from home or telecommute meant millions retained their jobs
and, to a varying extent, maintained productivity during periods of strict
lockdown around the world.
However, this ability has not been evenly distributed socially or spatially,
creating a new type of digital divide.
On one side are those who can work from home, supported by digital technologies, 
and have thus been able to enjoy both economic resilience and greater personal safety.
On the other side, previously employed individuals have been forced to accept
furlough or redundancy packages unless they are part of the cadre of essential
workers, who are potentially at high risk of infection.
Whilst the basis for this new digital divide can be viewed as mainly occupational, 
we consider whether the divide is also technological.
Using the UK as a case study, this paper aims to understand how the resilience
of broadband speeds in terms of both quality and reliability of service may reinforce
or redress the spatial and social dimensions of the digital division exposed by the
pandemic.
To do so, we employ volunteered geographic data on individual broadband speed tests
and state-of-the-art time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore how they overlap with or diverge from
the existing economic and digital geography of the UK. **need to add in urban-rural**
Our analysis enables us to better understand how the spatial and social distribution 
of both occupations and online accessibility intersect to enable or hinder the 
practice of telecommuting at a time of extreme demand. 
We will also consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a common means of accessing work, and broadband
services and infrastructure must be fit for purpose.
**LET'S LEAVE IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ**

<!-- Our framework: capability to WFH; HB old PARA2 -->
The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and are
allowed to be performed using digital technologies independently of 
location or co-location with colleagues, including supervisors 
[@allen2015effective; @singh2013modeling].
However, successful telecommuting also requires that the quality and
reliability of digital services, particularly home internet connection speeds,
enable the completion of work tasks with a minimum of delay or interruption.
Prior to the pandemic, the performance of broadband services with respect to
telecommuters was never tested at scale, as working from home and connecting to
colleagues and workplace resources via broadband was the purview of a small minority 
of workers. 
Instead, in the UK, for example, benchmarking broadband performance and service 
delivery by different Internet Service Providers (ISPs) focused upon the quality of 
service for evening leisure use, when video streaming services are at their peak, 
rather than during the working day [@ofcom2017].
Yet the shift towards telecommuting during various stages of lockdown around the world
has been drastic and there are speculations that post-Covid, the tendency to work from
home will be much higher than the pre-Covid, raising questions around whether 
internet services can accommodate the increased demand.
For example, $47$% of people in employment in the UK worked soley from home in April
$2020$, whilst the same figure only reached $5$% the year before [@ons2020; 
@ons2020lm2019]. 
A back of the envelope calculation suggests that up to 40% of the working force
could work from home on an ongoing basis [@batty2020editorial].
Similar figures have been reported for other countries around the world
[@felstead2020homeworking].
For instance, $37$% of the European workforce worked from home in April $2020$ with
countries like Finland reaching $60$% [@eurofound2020].
In the US, almost half of the working population worked from home during the same
period because of the pandemic [@brynjolfsson2020covid].
Furthermore, a recent estimate indicated that $37$% of all jobs in the US can be
permanently performed entirely from home [@NBERw26948].

<!-- ICT infrastructure and digital divides-->
None of these changes could have happened in the absence of reliable ICT
infrastructure -- both in terms of software and hardware.
But while software innovations are easily diffused across space and
society^[See for example the huge success of videoconferencing apps
[@marks2020zoom].],
the same does not apply for ICT hardware infrastructure such as internet
broadband connectivity.
The literature describes first level digital divides in terms of the availability and
quality of internet connectivity, such as that manifest in different geographies 
in the UK [@riddlesden2014broadband; @philip2017digital]. 
Second level digital divides consider the presence or lack of the necessary skills 
to effectively utilise digital technologies and the internet 
[@blank2014dimensions; @van2011internet].
The third level focuses on the differentiated returns of internet use **for...?**
[@stern2009levels; @van2014digital; @van2015third].
The capability to telecommute is related to all three levels of digital divides,
but more importantly leads to differentiated outcomes regarding the economic 
resilience of people and places during the current pandemic.
We identify this as a new digital divide, one that fundamentally alters the 
potential returns of internet use for the user and wider community, 
assumes skills that are present in some occupations but not in others, 
and relies upon access to high quality internet services.
As the quality of internet infrastructure and services, as well as variation in 
occupations are spatially dependent and clustered in space, our approach offers a 
framework for understanding which types of places, socially and spatially, 
are more likely to land on the right side of this new digital divide. 
By asking how resilient broadband speeds are as experienced in different parts of the 
UK during a time of extreme demand, we will better understand which places benefit 
from greater economic resilience not only during the pandemic, but also into the 
future.

The structure of this paper is as follows.
First we will review the literature on telecommuting and digital divides to better 
understand the origins of the new digital divide revealed by the pandemic and its
impact on the economic resilience of different places.
We then describe our data and methodology.
Our results section first describes the how internet services vary in the 
different clusters and then assesses whether these clusters replicate or repudiate
other socio-economic and geographic patterns of economic resilience.

...

@@

@@

<!-- Who works from home-->
In this analysis, the terms ‘telecommuting’ and ‘working from home’ are used 
interchangeably, as most remote labour during the Covid-19 crisis was carried out 
in the homes of individual employees rather than any other location [@eurofound2020]. 
However, it should be noted that previous research has explored how telecommuting 
can occur in other places, including satellite offices or on public transport 
[@felstead2012rapid; @siha2006telecommuting].
Previous research has also used a variety of definitions to measure the level of 
telecommuting within different workforces, distinguising, for example, between those 
directly employed, indirectly employed, self-employed, full-time or part-time, and 
who use digital technologies to work remotely full-days or part-days 
[@allen2015effective; @bailey2002review; @haddad2009examination].
<!--old text: An appropriate definition of teleworking is ‘the remote provision of labour that would otherwise be carried out within company premises’ (European Commission, 2020b).-->
No matter the definition, the option and capability to telecommute or work from home 
has never been equally distributed spatially or socio-economically any more than 
different industries and employment opportunities have.
For example, studies from the United States, the Netherlands, and the UK indicate 
that telecommuters are most likely to hold professional, managerial, and technical 
occupations, where the workforce is better educated and wealthier, and that there is
suppressed demand among women and part-time workers
[@headicar2016move; @peters2004employees; @singh2013modeling].
Opportunities for working from home during the current pandemic have likewise not 
been equally spread across the workforce.
@NBERw26948 indicated that in the US, managers, educators, as well as those
working in computer-related occupations, finance, and law can easily work from
home, and that occupations with opportunities to telecommute are associated
with higher earnings.
This is not the case for the workforce occupied in more spatially fixed occupations,
from farming, construction, and manufacturing to hospitality and care services.
In the US, these occupations tend to be lower-income, non-white, 
without a university degree, live in rental accommodation and lack health insurance
[@NBERw27085].
Similar trends can be observed for other countries.
For example, $75$% of workers with tertiary education worked from home in Europe
during spring $2020$, whilst only $34$% of workers with secondary education and
$14$% of those primary education did so [@eurofound2020].

## digital divides and economic resilience
<!-- WFH and digital divides-->
Our understanding of telecommuting as a product of enabled occupations can be 
described as a manifestation of the third level digital divide, as those who are able 
to use digital technologies to work from home benefit from a high rate of return on 
their use of the internet in terms of autonomy, flexibility, and time saved from 
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
These returns have been even greater during the Covid-19 crisis, when those with the 
ability to telecommute also have the ability to maintain their employment whilst 
protecting their health.
However, the success of these arrangements has also been dependent upon the first 
level digital divide, which is associated with access and quality of internet 
connectivity.
For example, the systematic review from @SALEMINK2017360 highlights the 
infrastructure quality differences between urban and rural areas.
Whether this variation in infrastructure quality affects the spatial footprint of 
telecommuting has not previously been measured, although there are indications that
those who purchase high speed connections consume more data of all sorts and use their
connections for a variety of purposes [@hauge2011consumer].
Another study identifies a correlation between access to internet services and a 
reduction in household transport spend across 33 coutries, although the authors are 
uncertain whether this implies less travel, for example because of increased
telecommuting, or whether internet access enables more efficient travel 
[@bris2017ict].
However, the extreme demand during the pandemic provides a new opportunity to 
understand how infrastructure accessibility, quality, and reliability affects 
telecommuting, particularly as working from home during the pandemic required
high volumes of bandwidth-intensive video conferencing in order to avoid the 
face-to-face contact that could increase the spread of infection.
Thus by answering questions about internet service resilience, we can also refine our
understanding of how this has contributed to or reduced the new digital divide where
economic resilience has been dependent upon the capability to work from home.

<!-- Econ resilience-->
This is aligned with the regional economic resilience literature, which
underlines the differentiated capacity of cities and regions to escape or
recover from economic crises [@martin2012regional; @kitsos2018economic].

## other subsection?

@@

@@

<!-- HB old PARA3 -->
This paper assesses the quality and reliability of upload speeds by time of day and 
day of the week across local authorities in the UK during the time when the population
were told to work from home if at all possible.
The experience of upload speeds is not the same as the maximum speeds offered by an 
ISP or possible speeds over a particular type and length of connection. 
We use volunteered geographic information in the form of speed-checks run by users to 
test their experienced broadband speeds, upload and download. 
Since the quality and reliability of internet services vary in time and space due to 
both supply and demand-side influences, we use a number of different measures of 
experienced upload speeds. 
These include: a) mean, experienced connection speed, 
b) standard deviation or the amount of fluctuation from the mean, and 
c) the variation in speeds at particular times of day when working from home is more 
likely to take place. 
We take account of all three measurements in order to describe upload speeds as fully 
as possible. 
We start by calculating the mean upload speed for every local government district in 
the UK for each hour of each weekday, excluding midnight to 6:00, weekends and bank 
holidays. 
We then cluster the local authorities by these temporal profiles, allowing us to 
identify patterns and describe the overall means, standard deviations, and other 
relevant statistics for each spatial cluster. 
Thus we aim to answer the first part of our research question: 
How resilient are broadband speeds as experienced in different parts of the UK during 
a time of extreme demand?

@@

# Conclusions
<!-- WFH broader consequences-->
The long-term effects of such drastic changes in telecommuting and attitudes
towards working from home are difficult to predict. Nevertheless,
they span through various aspects of economy and society: from changes to
transportation planning due to altered commuting patters, to changes in land
use and urban planning to accommodate people who work from home
[@BUDNITZ2020102713]; and from productivity and innovation changes, to changes
in agglomeration externalities and the attraction of large cities [@econobs]
just to name a few. This paper is positioned to support endeavours in
understanding the effects of increased telecommuting by
exposing the spatial and social dimensions of telecommuting including the
resilience of broadband speeds in terms of both quality and reliability of
service, and whether this reinforces or redresses prior digital divisions.
**HANNAH, HOW DOES THIS SOUND - this last bit probably in the discussion at the end?**
